\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2016-09-02 manuscript.tex

The key recommendation arising from our results is that researchers not use the LSM as an alternative to measuring potential confounding variables and/or explicitly modeling the endogenous network structure. The LSM cannot control for unmeasured confounders. We do not argue that this property is unique to LSM. Rather, we suspect it is the case with most, if not all, latent variable models (e.g., ideal point models, topic models, .

We conclude that the primary reason for using the LSM or one of its many variants, rather than a GLM, should be interest in using the latent space model without covariates to explore latent structure in the network through a principled approach to embedding, or the prediction of edges in networks where there is likely to be omitted structure that is not adequately modeled using observed covariates. Although inferential and estimation errors are somewhat smaller than that of a GLM when there is omitted network structure, even moderate collinearity between the omitted structure and the measured covariate leads to substantial bias and inferential error. The LSM cannot control for unmeasured confounders. Furthermore, if such network structure does not exist, the LSM may induce additional bias and inferential error, though this effect can be moderated by scaling the prior on the latent positions to match that of the measured covariates. If the researcher is interested in identifying causal effects of observed covariates, but is concerned that unmeasured variables or network structure could confound the relationship between observed variables and the network, the LSM does not represent an advisable alternative to measuring the confounding variables and/or explicitly modeling the endogenous network structure.

>>>>

The key recommendation arising from our results is that researchers not use the LSM as an alternative to measuring potential confounding variables and/or explicitly modeling the endogenous network structure. The LSM cannot control for unmeasured confounders. This property is, perhaps, not surprising. Given that all priors are centered at zero, the parameter regions that exhibit high posterior probability will be those that explain the observed network (i.e., high likelihood) using minimal departures from zero (i.e., high prior probability). Since explaining the observed network using a covariate requires moving one parameter value away from zero, whereas explaining the network using a latent dimension requires moving many latent coordinates away from zero, the LSM exhibits an inherent preference to explain network structure using observed covariates over latent coordinates. One way to address this issue would be to use a highly diffuse prior for the latent positions, but as we discuss above, that is not possible with the LSM. If the LSM is to be used as a tool for drawing inferences regarding observed covariate effects while adjusting for confounding network structure, further development is needed to enable the LSM to avoid bias and inferential error under confounding.
